Long-term perspectives of social inequality in mortality among

### 2017-07-03 article.Rmd
more translation from Rnw

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
library(knitr)
library(eha)
library(xtable)
library(skum)
```

```{r readalldata}
## Load remote data:
##con <- url("http://capa.ddb.umu.se/data/vb0.rda")
load("data/vb0.rda")
##load(con)
##close(con)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##
##con <- url("http://capa.ddb.umu.se/data/vb.rda")
load("data/vb.rda")
##load(con)
##close(con)
##
##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
load("data/lisaold.rda")
##load(con)
##close(con)
##
##con <- url("http://capa.ddb.umu.se/data/loTab.rda")
##load(con)
##close(con)
load("data/loTab.rda")
loTab$period <- factor(loTab$period)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
```

@@

One of the central aspects of survival is social class and access to economic
and other resources. Even in present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death" \citep{liph95}. For a long time,
the general view has been that socio-economic health inequalities were high in
historical societies, probably higher than in modern societies. This is a
reasonable assumption since these societies were in most cases characterised by
very large differences between groups and classes. Knowing that access to
resources provides advantages in all ways, the health advantage of higher
classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social
inequality in health has passed through different historical phases. According
to them, social health differences were comparatively small during the
pre-transitional phase when space was a strong determinant for the spread of
disease. Differences then increased during the transitional phase when mortality
declined and wealthy groups used their resources to gain better health. Finally,
health differences decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal differences.
\citet{omran82} state, in the third proposition in his theory of the
epidemiological transition, that even if the class differential in mortality was
maintained during the transition, the decline set in earlier and was faster
among privileged groups.

Recent studies, investigating social inequality in health and mortality
with micro-data, has however questioned how generalizable the assumed
pattern is \citep{bevp11}. Instead, some studies suggest that the modern
pattern might be a fairly recent development, although the pattern differs
depending on age, gender or historical context. When it comes to the adult
and elderly population results from different studies have diverged, some
reporting substantial differences but many, particularly those based on
Swedish data, have found surprisingly small or even reversed socio-economic
differences during the first stages of the transition \citep{bedr11,
edli11,edbr12}, sharply contrasting against what we know from present-day
society.

Another question of the development of social inequality in mortality,
concerns its relation to economic inequality. \citet{wipi09} argue that
income inequality has an independent effect on mortality, separate from the
direct effect of actual different access to economic resources. They argue
that unequal societies perform less well when it comes to health than equal
ones. Wilkinson and Pickett find a positive association between inequality
and health (as well as other social conditions) at country level in the
present-day economically developed world. This has initiated a vital
scholarly debate and the topic has been extensively studied \citep{suka04,
wado00}. How the association looked like in previous periods is
unknown. This is of particular interest since the levels of economic
inequality were very different from those of the recent decades. Even if it
not a necessary implication, we would still assume that poorer groups were
those most disadvantaged of large inequality in itself. The situation
during the 19th century is uncertain, but at least during the late 19th and
early 20th century both income and wealth distribution was strongly skewed
in Sweden. From a couple of decades into the 20th century, income and
wealth inequality continuously diminished to reach a low point at around
1980. Thereafter, economic inequality has increased substantially until the
present, although far from being as high as a century ago \citep{rowa08,
rowa09}. The Swedish development resembles that of other European and North
American countries in its basic features \citep{rowa14}.

@@

The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the
county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf
of Bothnia. This was a remote part of the country, where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures, and
several severe famines affected the regions during the 1800s. One living
long in memory was the famine of 1867–-1868 \citep{segb14}. During the long
winters, sea communication was hindered due to the Gulf of Bothnia being
frozen. The first ships during the year came to this area in May or even as
late as June in some cases \citep{fah56}. Towards the end of the 19th
century, the Swedish railway system reached this part of Sweden, thereby
facilitating contacts with the rest of Sweden, improving the economy and
mitigating the effects of harvest failures. The inland was particularly
remote and isolated. During the 19th century, our regions became
increasingly integrated in the same epidemiological pattern as the rest of
Sweden. Diseases reached to places that earlier were spared.

@@

Mortality was fairly low in comparison with other parts of the country and
the fertility transition was late \citep{coa86}. The Skellefteå population
size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on
January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the
century.

@@

The recent extension in time of data at the Demographic Data Base (DDB),
Umeå University (www.cedar.umu.se) by including the period between 1900 and
1950s makes it possible to investigate the long-term development of social
inequality in mortality, in the present case from 1851 to 2013. The
databases provide us with micro-data on a sufficiently large population and
with substantial social diversity for the Skellefteå and Umeå regions
\citep{wee16}. The data for the present study comes from two large
population databases at the DDB. The early period until 1950 is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked
parish records, allowing us to reconstruct life biographies on people as
long as they remained in the region. The records are linked within but not
between the regions. Data from Poplink are accessed for the period 1851--1950.

The other large data-set is extracted from the Linnaeus database
\citep{malm10}. Linnaeus is based on different linked national population
registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from
censuses every fifth year 1960--1990, with additional information on deaths
from National Board of Health and Welfare. For the period 2002--2013 we use
the information from the yearly population registers from Statistics Sweden
(LISA) together with death information from National Board of Health and
Welfare.

@@

In the data-set analysed here, all individuals 40 years and older ever
being resident in either of the regions are included. The data file
contains the variables on social class ({\sc Hisclass}) sex, urban/rural
residence, birth date, death date, first and last date of observation and
type of entrance/exit. Total number of persons are 84\,599 leading to
35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in
the Linnaeus database. See Figure~\ref{fig:lexis2}.

@@

In this paper we analyse mortality differences according to social class
based on the classification scheme {\sc Hisclass} \citep{lema11}. The only
type of socio-economic information available both in Linnaeus and Poplink
is occupations. Thus, analysing according to social class is apart from
being a scientific choice, also a practical one. Using occupational data is
the only and obvious alternative for making a longitudinal study of
socio-economic mortality differences with consistent classification
schemes. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between
sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the
original DDB coding system. The censuses 1960--1990 include occupational
codes based on the system used in Sweden at that time, NYK (Nordisk
Yrkesklassificering). The yearly population registers (LISA) do not report
occupations until 2001, thus there is a window of 11 years from 1991 to
2001 where we lack information. In the censuses, we have solved this by
extending the periods for which we consider the observed social class to be
valid. Thus, the social class in the census 1990 is considered to be kept
until 2001. This strategy is also used for other cases of missing social
class. Furthermore, we consider the last social position before retirement
to be kept until death or last observation for those above 65. For the
Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided
to us by Jonas Helgertz, Lund University.

The use of different data sets, forces us to apply different approaches
towards defining the social position in some respects. A problem is that
Poplink usually only provides the occupation for the head of household,
thus under-estimating female working-force participation as well as that of
adult children residing with their parents \citep{vik10}. In this study and
according to the research questions in the present paper, we have chosen to
categorise wife according to the position of the head of household, usually
the husband/father, assuming that the family shares the same socio-economic
position. This strategy makes the problem of under-recording less
problematic.

Female labour is much better covered in the Linnaeus database. A
complication is however that it is  difficult to define households in the
same way as in Poplink. Extra-marital cohabitation, connected to the Second
Demographic Transition \citep{van87, van03}, became common and female
labour force participation developed as the norm in Sweden during this
period. The occupation of the husband as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable and makes the
interpretation of the development for women more shaky.

{\sc Hisclass}, the classification system used here, is a "\ldots
HISCO-based historical international social class scheme" \citep{lema11,
lema02}. The different classes in {\sc Hisclass} represent distinct
categories, based on whether the work is manual or non–manual, skill level,
supervision and sector. Implicitly it reflects large differences in what is
in the forefront of other categorisations, as for example access to
economic and other resources, status, social power et cetera. We have
merged the original 12 classes in {\sc Hisclass}, based on position in the
labour market, ending up with the following three and an additional fourth
category with unknown class:

\begin{enumerate}
\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals
and higher white collar.
\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers,
farmers and lower white collar.
\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including
farm workers.
\end{enumerate}

Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution
(per cent) of exposure time according to class for men and women
respectively. The category of workers is fairly stable over time. The
highest class has increased substantially, although from a very low level
in the 19th century. The increase of academics in connection to the
establishment of Umeå University explains much of this increase. What is
hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle
group from a majority of farmers during the 19th century to lower white
collar from the middle of the 20th century. Furthermore we observe an
increase during the 1960s and partly during later decades as well in those
with undefined. When first observed in the earliest included censuses,
occupation is often missing. For men the proportion is somewhat higher in
the age group 65 and above, i.e. those that were retired, but still low
enough to be used in our analyses. Until 1950, there were however very few
with missing occupation. For men in working-active age, the proportion
missing does not cause any problems for our analyses.

@@

The proportion with missing values among women is however a larger problem,
see Figure~\ref{fig:hiscskumw}. Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century. This is a reflection of how womens social position is defined for
that period. Missing social class is however a real problem during the
1960s and 1970s in particular. Around 80 \% lack information for the 1960s
and 50 \% for the 1970s. Since social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce, the proportion missing becomes very high. We have chosen not
to report results for women for these two decades. For the later decades,
the results are fairly complete and should be representative.

@@

Regarding the distribution of marital status over time, see
Figure~\ref{fig:civstskum}. The figure indicates an under-registration of
those married and and over-registration of previously married during the
1960s.

@@

The distribution of causes of death is shown in
Figure~\ref{fig:causeskum}. Notice the high proportion of other causes in
the first two periods. This category include cases with unknown or
unspecific causes which has implications for the cause-specific analyses
(see later discussion). The total dominance of our two disease groups is
apparent.

@@

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for civil status and rural/urban environment. Since
the central explanatory variable, {\sc Hisclass}, do not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

Another practical problem is the size of that data set:
It contains no less than 2432871 records of 236011 individuals, of
which 68244 are observed to die.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three civil statuses, ten periods, two
urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.

@@

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha}
\citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were
retired. The model controls for marital
status and whether the individual resides in the urban or rural parts of the regions. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

@@

# The effect of {\sc Hisclass}, 1851--1950, 1961--2013

@@

Figure~\ref{fig:cumsw} shows the summed hazards of dying in the age interval
40--64 during the different periods. Obviously, and not surprisingly, mortality
is decreasing over time for both sexes. Women have consistently better survival
than men. This is a well-known phenomenon in Swedish
demographic history. Male mortality has almost always been higher than
female in all age groups \citep{wil99, suwi07}. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. The compression that is
visible for later decades is partly explained by the decreasing general
mortality and is not necessarily representing a homogenisation of relative risks.

By \emph{standardising} the levels in each time period for each class, we illustrate the relative risks in Figure~\ref{fig:cumswstand}. In the rest of the paper, we only present these
relative mortality levels for the different classes.

@@

The most striking result of the analysis is the different patterns for men
and women. The social pattern of mortality among women is quite much in
accordance with previous analyses from the present research group on other
regions in northern Sweden and for other periods \citep{edli11,edbr12}. The
highest social class usually have comparatively low mortality, while
working class women have high. There is however no substantial change in
the relative risks between the classes. Only the group with unknown social
class is distinguished by moch higher mortality from the 1980s
onwards. This can however have to do with the differences how social
position for women is defined before and after 1950. After 1950 it is based
on the own occupation. Being without occupation might include women that
are more frail both when it comes to health and socially.

A different pattern appears among men, being the most surprising finding in
our analyses. The groups we would expect to have the lowest mortality, if
we assume that access to different resources and having a high status
determines survival, instead have by far the highest hazards of dying
during the nineteenth and much of the 20th century. This is also in
accordance with our previous studies \citep{edli11, edbr12}. Workers on the
other hand have comparatively low mortality until the 1960s, at the same
level as the middle class. The relative positions change radically towards
the end the 20th century. From the 1980s, there is a clear advantage when
it comes to survival for higher managers and professionals. The relative
differences are however more compressed than for the period until 1950. The
expected social gradient in mortality has developed. This is also what we
usually find in analyses of social health inequalities in present-day
Sweden as well as in most other countries. In the regions we are studying,
this is however a quite recent phenomenon. Corresponding to what we find
for women, those without known occupation have much higher mortality than
the others from the 1960s onwards. The assumption made for women of
unknowns social class is probably even more relevant for men. They probably
represent groups in society that are particularly vulnerable.

A further observation can be made from Figure~\ref{fig:cumsw}. The gender
difference in death hazards is very big for the élite group from
1851--1950, men having about three times higher risks than women in the
same class. In the other classes, the difference was much smaller. It was
men in high position that experienced the largest risks.

@@

Concerning the social pattern, the pattern is less clear compared to what
we found for the younger age group (see Figure~\ref{fig:cumswold}). For
women the results are mixed. Once again, too much missing data on social
class leaves out the results for the 1960s and 1970s. We do not find the
same clear advantage for élite women during period until 1950, but neither
do we find any large disadvantage for working class elderly women at that
time. In some periods, the social differences were small, and the élite
even had substantially higher mortality during the early 20th century. The
social differences continued to be small during the last decades of the
20th century. Only during the two last periods (after the year 2000) is it
possible to establish the expected social gradient in mortality.

@@

Performing the same types of analyses as for all-cause mortality, we start
by looking at cumulative hazards by {\sc Hisclass} at age 65, given
survival to age 40.

@@

Figure~\ref{fig:41cumsw} shows mortality levels during different periods
from 1901 on-wards. Women from the highest social class had consistently
lower mortality from cardiovascular diseases compared to the others. We
also observe increasing social differences from the first half of the
century to the last decades. A new pattern has developed from 1980
onwards. Now working class women have substantially higher cardiovascular
mortality compared to others. We can however not exclude that this is an
artefact of different diagnostics for deaths in different groups. Working
class women may have had less precise diagnoses in the early 20th century.

@@

See Table~\ref{tab:12cumswtab}.

>>>>

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(engine='R',fig.path='figs/art-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE)##$
library(knitr)
library(eha)
library(xtable)
library(skum)
```

```{r readalldata}
source("R/get.file.R")
vb0 <- get.file("vb0")
##if (file.exists("data/vb0.rda")){
##    load("data/vb0.rda")
##}else{
## Load remote data:
##   con <- url("http://capa.ddb.umu.se/data/vb0.rda")
##   load(con)
##   close(con)
##   save(vb0, file = "data/vb0.rda")
##}
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##

vb <- get.file("vb")
##if (file.exists("data/vb.rda"))
##con <- url("http://capa.ddb.umu.se/data/vb.rda")
##load("data/vb.rda")
##load(con)
##close(con)
##
##con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
lisaold <- get.file("lisaold")
##load("data/lisaold.rda")
##load(con)
##close(con)
##
##con <- url("http://capa.ddb.umu.se/data/loTab.rda")
##load(con)
##close(con)
loTab <- get.file("loTab")
##load("data/loTab.rda")
loTab$period <- factor(loTab$period)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
```

@@

One of the central aspects of survival is social class and access to economic
and other resources. Even in present-day welfare societies, social position is a
strong determinant when it comes to health and mortality and the impact even
seems to be increasing [@kuetal04, @frlu07]. It has been suggested that
"... social conditions have been, are and will continue to be irreducible
determinants of health outcomes and therefore deserves their appellation as
'fundamental causes' of disease and death" [@liph95]. For a long time,
the general view has been that socio-economic health inequalities were high in
historical societies, probably higher than in modern societies. This is a
reasonable assumption since these societies were in most cases characterised by
very large differences between groups and classes. Knowing that access to
resources provides advantages in all ways, the health advantage of higher
classes ought to be obvious. @ant67 and @smi83 suggest that social
inequality in health has passed through different historical phases. According
to them, social health differences were comparatively small during the
pre-transitional phase when space was a strong determinant for the spread of
disease. Differences then increased during the transitional phase when mortality
declined and wealthy groups used their resources to gain better health. Finally,
health differences decreased again in modern low-mortality settings when instead
health-related behaviour became important, resulting in marginal differences.
@omran82 states, in the third proposition in his theory of the
epidemiological transition, that even if the class differential in mortality was
maintained during the transition, the decline set in earlier and was faster
among privileged groups.

Recent studies, investigating social inequality in health and mortality
with micro-data, has however questioned how generalizable the assumed
pattern is [@bevp11]. Instead, some studies suggest that the modern
pattern might be a fairly recent development, although the pattern differs
depending on age, gender or historical context. When it comes to the adult
and elderly population results from different studies have diverged, some
reporting substantial differences but many, particularly those based on
Swedish data, have found surprisingly small or even reversed socio-economic
differences during the first stages of the transition [@bedr11,
@edli11,@edbr12], sharply contrasting against what we know from present-day
society.

Another question of the development of social inequality in mortality,
concerns its relation to economic inequality. @wipi09 argue that
income inequality has an independent effect on mortality, separate from the
direct effect of actual different access to economic resources. They argue
that unequal societies perform less well when it comes to health than equal
ones. Wilkinson and Pickett find a positive association between inequality
and health (as well as other social conditions) at country level in the
present-day economically developed world. This has initiated a vital
scholarly debate and the topic has been extensively studied [@suka04,
@wado00]. How the association looked like in previous periods is
unknown. This is of particular interest since the levels of economic
inequality were very different from those of the recent decades. Even if it
not a necessary implication, we would still assume that poorer groups were
those most disadvantaged of large inequality in itself. The situation
during the 19th century is uncertain, but at least during the late 19th and
early 20th century both income and wealth distribution was strongly skewed
in Sweden. From a couple of decades into the 20th century, income and
wealth inequality continuously diminished to reach a low point at around
1980. Thereafter, economic inequality has increased substantially until the
present, although far from being as high as a century ago [@rowa08,
@rowa09]. The Swedish development resembles that of other European and North
American countries in its basic features [@rowa14].

@@

```{r mapp,fig.cap = "The regions."}
include_graphics("myfigs/sweden.png")
```

The Skellefteå and Umeå regions (Figure \@ref(fig:mapp)) are part of the
county of Västerbotten in the north of Sweden along the coast of Gulf
of Bothnia. This was a remote part of the country, where communication with
the rest of Sweden was difficult until the late 19th century. The economy
was dominated by agriculture, making it vulnerable to harvest failures, and
several severe famines affected the regions during the 1800s. One living
long in memory was the famine of 1867–-1868 [@segb14]. During the long
winters, sea communication was hindered due to the Gulf of Bothnia being
frozen. The first ships during the year came to this area in May or even as
late as June in some cases [@fah56]. Towards the end of the 19th
century, the Swedish railway system reached this part of Sweden, thereby
facilitating contacts with the rest of Sweden, improving the economy and
mitigating the effects of harvest failures. The inland was particularly
remote and isolated. During the 19th century, our regions became
increasingly integrated in the same epidemiological pattern as the rest of
Sweden. Diseases reached to places that earlier were spared.

@@

Mortality was fairly low in comparison with other parts of the country and
the fertility transition was late [@coa86]. The Skellefteå population
size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on
January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the
century.

@@

The recent extension in time of data at the Demographic Data Base (DDB),
Umeå University (www.cedar.umu.se) by including the period between 1900 and
1950s makes it possible to investigate the long-term development of social
inequality in mortality, in the present case from 1851 to 2013. The
databases provide us with micro-data on a sufficiently large population and
with substantial social diversity for the Skellefteå and Umeå regions
[@wee16]. The data for the present study comes from two large
population databases at the DDB. The early period until 1950 is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region [@wee16]. Poplink is based on linked
parish records, allowing us to reconstruct life biographies on people as
long as they remained in the region. The records are linked within but not
between the regions. Data from Poplink are accessed for the period 1851--1950.

The other large data-set is extracted from the Linnaeus database
[@malm10]. Linnaeus is based on different linked national population
registers from 1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from
censuses every fifth year 1960--1990, with additional information on deaths
from National Board of Health and Welfare. For the period 2002--2013 we use
the information from the yearly population registers from Statistics Sweden
(LISA) together with death information from National Board of Health and
Welfare.

@@

In the data-set analysed here, all individuals 40 years and older ever
being resident in either of the regions are included. The data file
contains the variables on social class (Hisclass) sex, urban/rural
residence, birth date, death date, first and last date of observation and
type of entrance/exit. Total number of persons are 84\,599 leading to
35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in
the Linnaeus database. See Figure @\ref(fig:lexis2).

@@

In this paper we analyse mortality differences according to social class
based on the classification scheme Hisclass [@lema11]. The only
type of socio-economic information available both in Linnaeus and Poplink
is occupations. Thus, analysing according to social class is apart from
being a scientific choice, also a practical one. Using occupational data is
the only and obvious alternative for making a longitudinal study of
socio-economic mortality differences with consistent classification
schemes. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between
sources. Occupations in Poplink are coded into HISCO [@lema02] and the
original DDB coding system. The censuses 1960--1990 include occupational
codes based on the system used in Sweden at that time, NYK (Nordisk
Yrkesklassificering). The yearly population registers (LISA) do not report
occupations until 2001, thus there is a window of 11 years from 1991 to
2001 where we lack information. In the censuses, we have solved this by
extending the periods for which we consider the observed social class to be
valid. Thus, the social class in the census 1990 is considered to be kept
until 2001. This strategy is also used for other cases of missing social
class. Furthermore, we consider the last social position before retirement
to be kept until death or last observation for those above 65. For the
Linnaeus data, we use a script recoding HISCO into Hisclass, provided
to us by Jonas Helgertz, Lund University.

The use of different data sets, forces us to apply different approaches
towards defining the social position in some respects. A problem is that
Poplink usually only provides the occupation for the head of household,
thus under-estimating female working-force participation as well as that of
adult children residing with their parents [@vik10]. In this study and
according to the research questions in the present paper, we have chosen to
categorise wife according to the position of the head of household, usually
the husband/father, assuming that the family shares the same socio-economic
position. This strategy makes the problem of under-recording less
problematic.

Female labour is much better covered in the Linnaeus database. A
complication is however that it is  difficult to define households in the
same way as in Poplink. Extra-marital cohabitation, connected to the Second
Demographic Transition [@van87, @van03], became common and female
labour force participation developed as the norm in Sweden during this
period. The occupation of the husband as signifier of social class became,
if not obsolete, at least less relevant. All included persons are signified
by their individual occupation for this period. Still, the results for the
different periods are not completely comparable and makes the
interpretation of the development for women more shaky.

Hisclass, the classification system used here, is a "\ldots
HISCO-based historical international social class scheme" [@lema11,
@lema02]. The different classes in Hisclass represent distinct
categories, based on whether the work is manual or non–manual, skill level,
supervision and sector. Implicitly it reflects large differences in what is
in the forefront of other categorisations, as for example access to
economic and other resources, status, social power et cetera. We have
merged the original 12 classes in Hisclass, based on position in the
labour market, ending up with the following three and an additional fourth
category with unknown class:

\begin{enumerate}
\item {\tt elite}, Hisclass 1 and 2. Higher managers, professionals
and higher white collar.
\item {\tt middle}, Hisclass 3, 4, 5, 6 and 8. Lower managers,
farmers and lower white collar.
\item {\tt worker}, Hisclass 7, 9, 10, 11 and 12. Workers including
farm workers.
\end{enumerate}

Figures @\ref(fig:hiscskumm) and @\ref{fig:hiscskumw} show the distribution
(per cent) of exposure time according to class for men and women
respectively. The category of workers is fairly stable over time. The
highest class has increased substantially, although from a very low level
in the 19th century. The increase of academics in connection to the
establishment of Umeå University explains much of this increase. What is
hidden in Figure @\ref(fig:hiscskumm) is the transfer within the middle
group from a majority of farmers during the 19th century to lower white
collar from the middle of the 20th century. Furthermore we observe an
increase during the 1960s and partly during later decades as well in those
with undefined. When first observed in the earliest included censuses,
occupation is often missing. For men the proportion is somewhat higher in
the age group 65 and above, i.e. those that were retired, but still low
enough to be used in our analyses. Until 1950, there were however very few
with missing occupation. For men in working-active age, the proportion
missing does not cause any problems for our analyses.

@@

The proportion with missing values among women is however a larger problem,
see Figure @\ref(fig:hiscskumw). Corresponding to what we found for men, the
proportion without social position is very low until the middle of the 20th
century. This is a reflection of how womens social position is defined for
that period. Missing social class is however a real problem during the
1960s and 1970s in particular. Around 80 \% lack information for the 1960s
and 50 \% for the 1970s. Since social class for this period is defined
according to own occupation, the consequence of many women being outside
the workforce, the proportion missing becomes very high. We have chosen not
to report results for women for these two decades. For the later decades,
the results are fairly complete and should be representative.

@@

Regarding the distribution of marital status over time, see
Figure \@ref(fig:civstskum). The figure indicates an under-registration of
those married and and over-registration of previously married during the
1960s.

@@

The distribution of causes of death is shown in
Figure \@ref(fig:causeskum). Notice the high proportion of other causes in
the first two periods. This category include cases with unknown or
unspecific causes which has implications for the cause-specific analyses
(see later discussion). The total dominance of our two disease groups is
apparent.

@@

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for civil status and rural/urban environment. Since
the central explanatory variable, Hisclass, do not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

Another practical problem is the size of that data set:
It contains no less than 2432871 records of 236011 individuals, of
which 68244 are observed to die.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} [@cox72,@gb02]. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} [@fisher22]. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three civil statuses, ten periods, two
urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.

@@

The analyses are performed in the \R\ environment for statistical computing
and graphics [@cran], especially using the package {\tt eha}
[@gb12,@eha]
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were
retired. The model controls for marital
status and whether the individual resides in the urban or rural parts of the regions. There
are no serious signs of deviation from proportionality assumption, except
regarding Hisclass. Each period are analysed separately, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

@@

# The effect of Hisclass, 1851--1950, 1961--2013

@@

Figure \@ref(fig:cumsw) shows the summed hazards of dying in the age interval
40--64 during the different periods. Obviously, and not surprisingly, mortality
is decreasing over time for both sexes. Women have consistently better survival
than men. This is a well-known phenomenon in Swedish
demographic history. Male mortality has almost always been higher than
female in all age groups [@wil99, @suwi07]. We also observe a
substantial decline in mortality for both sexes, resulting in smaller
absolute differences between social classes. The compression that is
visible for later decades is partly explained by the decreasing general
mortality and is not necessarily representing a homogenisation of relative risks.

By \emph{standardising} the levels in each time period for each class, we 
illustrate the relative risks in Figure \@ref(fig:cumswstand). In the rest of 
the paper, we only present these
relative mortality levels for the different classes.

@@

The most striking result of the analysis is the different patterns for men
and women. The social pattern of mortality among women is quite much in
accordance with previous analyses from the present research group on other
regions in northern Sweden and for other periods [@edli11,@edbr12]. The
highest social class usually have comparatively low mortality, while
working class women have high. There is however no substantial change in
the relative risks between the classes. Only the group with unknown social
class is distinguished by moch higher mortality from the 1980s
onwards. This can however have to do with the differences how social
position for women is defined before and after 1950. After 1950 it is based
on the own occupation. Being without occupation might include women that
are more frail both when it comes to health and socially.

A different pattern appears among men, being the most surprising finding in
our analyses. The groups we would expect to have the lowest mortality, if
we assume that access to different resources and having a high status
determines survival, instead have by far the highest hazards of dying
during the nineteenth and much of the 20th century. This is also in
accordance with our previous studies [@edli11, @edbr12]. Workers on the
other hand have comparatively low mortality until the 1960s, at the same
level as the middle class. The relative positions change radically towards
the end the 20th century. From the 1980s, there is a clear advantage when
it comes to survival for higher managers and professionals. The relative
differences are however more compressed than for the period until 1950. The
expected social gradient in mortality has developed. This is also what we
usually find in analyses of social health inequalities in present-day
Sweden as well as in most other countries. In the regions we are studying,
this is however a quite recent phenomenon. Corresponding to what we find
for women, those without known occupation have much higher mortality than
the others from the 1960s onwards. The assumption made for women of
unknowns social class is probably even more relevant for men. They probably
represent groups in society that are particularly vulnerable.

A further observation can be made from Figure \@ref(fig:cumsw). The gender
difference in death hazards is very big for the élite group from
1851--1950, men having about three times higher risks than women in the
same class. In the other classes, the difference was much smaller. It was
men in high position that experienced the largest risks.

@@

Concerning the social pattern, the pattern is less clear compared to what
we found for the younger age group (see Figure \@ref(fig:cumswold). For
women the results are mixed. Once again, too much missing data on social
class leaves out the results for the 1960s and 1970s. We do not find the
same clear advantage for élite women during period until 1950, but neither
do we find any large disadvantage for working class elderly women at that
time. In some periods, the social differences were small, and the élite
even had substantially higher mortality during the early 20th century. The
social differences continued to be small during the last decades of the
20th century. Only during the two last periods (after the year 2000) is it
possible to establish the expected social gradient in mortality.

@@

Performing the same types of analyses as for all-cause mortality, we start
by looking at cumulative hazards by Hisclass at age 65, given
survival to age 40.

@@

Figure \@ref(fig:41cumsw) shows mortality levels during different periods
from 1901 on-wards. Women from the highest social class had consistently
lower mortality from cardiovascular diseases compared to the others. We
also observe increasing social differences from the first half of the
century to the last decades. A new pattern has developed from 1980
onwards. Now working class women have substantially higher cardiovascular
mortality compared to others. We can however not exclude that this is an
artefact of different diagnostics for deaths in different groups. Working
class women may have had less precise diagnoses in the early 20th century.

@@

See Table \@ref(tab:12cumswtab).

Figure \@ref(fig:12cumsw) shows relative mortality levels according to
Hisclass during the different studied periods for the age group
40--64. Cancer mortality among the élite women was lower than that for the
other classes during the complete studied period, while working class women
had the highest levels during almost all periods. Consequently, we find a
more consistent advantage of higher class in cancers compared to
cardiovascular diseases for women. We do however not find the same pattern
for men. It is difficult to identify any general features of 20th century
male cancer mortality when it comes to social differences.
